Driver coverage

### 2024-02-05 infercnv/reports/infercnv.Rmd
Recover repository after accidental deletion of .git directory

```{r, results = 'asis'}
pngs <-
  list.files('out/Zhang2023/', pattern = '^BRI') %>% head %>%
  purrr::set_names() %>%
  purrr::map(function(sample_id) {
    list.files(paste0('out/Zhang2023/', sample_id), pattern = '.png', 
                     recursive = T, full.names = T) %>%
      {setNames(., basename(.))}
  })

@@

  chunks <- c(chunks, paste0('\n#### ', sample_id))

  # for (i in names(pngs[[sample_id]])) {
  #   
  #   chunks <- c(chunks, paste0('\n##### ', i))

    knit_plot <- c(
      '',
      '```{r, warning = F, message = F, echo = F}',
      paste0('slickR(obj = pngs[["', sample_id, '"]]) %synch% ',
             '( slickR(names(pngs[["', sample_id, '"]]), slideType = "p") + settings(arrows = F) )'),
      '```',
      ''
    )
    knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
    dev_off_if()

  }

  knit_plot <- c(
    '',
    '```{r, warning = F, message = F, echo = F}',
    paste0('slickR(obj = pngs[["', sample_id, '"]]) %synch% ',
           '( slickR(names(pngs[["', sample_id, '"]]), slideType = "p") + settings(arrows = F) )'),
    '```',
    ''
  )
  knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
  dev_off_if()

@@

}

>>>>

```{r, results = 'asis'}
pngs <-
  list.files('out/Zhang2023/', pattern = '^BRI') %>%
  purrr::set_names() %>%
  purrr::map(function(sample_id) {
    lis <-
      list.files(paste0('out/Zhang2023/', sample_id), pattern = '.png', 
                 recursive = T, full.names = T) %>%
      {setNames(., gsub('infercnv.', '', tools::file_path_sans_ext(basename(.))))}

    # reorder
    c(lis['infercnv'], lis[names(lis) != 'infercnv'])
  })

@@

  chunks <- c(chunks, paste0('\n### ', sample_id))

  for (i in names(pngs[[sample_id]])) {

    chunks <- c(chunks, paste0('\n#### ', i))

    knit_plot <- c(
      '',
      '```{r, warning = F, message = F, echo = F}',
      paste0('knitr::include_graphics("../', pngs[[sample_id]][[i]], '")'),
      '```',
      ''
    )
    knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
    dev_off_if()

@@

  }
}

### 2024-02-05 scomatic/reports/driver_coverage.Rmd
Recover repository after accidental deletion of .git directory

---
title: "Mutation loads in rheumatoid arthritis data from SComatic"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
    rerun: true
---

@@

knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = 'reports/driver_coverage_cache/')

@@

### Calculate coverage, by sample and by sample-x-celltype

We use `samtools depth` to get the coverage of each site in the drivers, both across each sample and within each cell type in each sample.

```{bash cov, eval = T}
# activate conda env
mamba activate jupy

# dirs
input_dir=/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/cellranger_output
ct_input_dir=out/Zhang2023/default_thresholds/
output_dir=out/Zhang2023/coverage/
drivers=data/driver_genes/driver_gene_coords_for_coverage.tsv
mkdir -p $output_dir

# iterate over each bam file and pileup specific regions
(
  echo -e 'chr\tpos\tdepth\tgene\tsample' ;
  for bam in $(find $input_dir -iname 'possorted_genome_bam.bam'); do
  	sample=$(echo $bam | cut -d "/" -f 10 | sed 's/\-/_/g')
    while read -r gene coords ; do
      samtools depth -a -r $coords -Q 30 $bam |
      awk -v sample=$sample -v gene=$gene '{print $0"\t"gene"\t"sample}'
    done < <(sed 1d $drivers) ;
  done
) | cat > $output_dir/drivers_samtools_depth.tsv

# iterate over celltype-specific bam files and pileup specific regions
(
  echo -e 'chr\tpos\tdepth\tgene\tsample\tcelltype' ;
  for sample_dir in $ct_input_dir/*-GEX ; do
    sample=$(basename $sample_dir | sed 's/\-GEX$//g')
    for ct_bam in $sample_dir/celltype_bams/*.bam ; do
      ct=$(basename $ct_bam | cut -f2 -d.)
      while read -r gene coords ; do
        samtools depth -a -r $coords -Q 30 $ct_bam |
        awk -v sample=$sample -v gene=$gene -v ct=$ct \
        '{print $0"\t"gene"\t"sample"\t"ct}'
      done < <(sed 1d $drivers) ;
    done
  done
) | cat > $output_dir/drivers_samtools_depth_by_celltype.tsv
```

We read in the results at both levels.

```{r read_cov}
cov <- xfun::cache_rds({
  cov_files <-
    list.files(
      'out/Zhang2023/coverage/', 
      pattern = 'drivers_samtools_depth.*.tsv',
      full.names = T) 
  cov <-
    cov_files %>%
    setNames(cov_files %>% basename() %>% tools::file_path_sans_ext()) %>%
    purrr::map(readr::read_tsv) %>%
    dplyr::bind_rows(.id = 'lvl') %>%
    dplyr::mutate(chr = as.character(chr))
  cov
}, file = 'cov.rds', rerun = params$rerun)
```

@@

```{r features}
# annotate gene features
cov <- xfun::cache_rds({

  # load rda to get ref_cds object from DNDSCV
  refcds_38 <- '/lustre/scratch125/casm/team268im/fa8/119/DNDSCV_COVARIATES/refcds_GRCh38-GencodeV18+Appris.rda'
  load(refcds_38)

  # set gene names
  ref_cds <-
    RefCDS %>%
    setNames(RefCDS %>% purrr::map(~ .x$gene_name) %>% unlist())

  # annotate exons
  exons <-
    ref_cds[c(drivers$gene)] %>%
    purrr::map(~ .x$intervals_cds %>% 
                 tibble::as_tibble() %>%
                 dplyr::transmute(chr = .x$chr, start = V1, end = V2) %>%
                 dplyr::mutate(exon = paste0('exon', dplyr::row_number()))) %>%
    dplyr::bind_rows(.id = 'gene') %>%
    dplyr::mutate(pos = purrr::map2(start, end, ~ .x:.y)) %>%
    tidyr::unnest(cols = 'pos') %>%
    dplyr::select(-start, -end)
  cov <-
    cov %>%
    dplyr::left_join(exons, by = c('chr', 'pos', 'gene')) %>%
    dplyr::mutate(exon = ifelse(is.na(exon), 'intron', exon),
                  feature = ifelse(exon == 'intron', 'intron', 'exon'))

@@

  cov %>% 
    dplyr::group_by(gene, sample) %>%
    dplyr::group_split() %>%
    purrr::map(function(df_i) {

      # for testing:
      # cov %>% dplyr::group_by(gene, sample) %>% dplyr::group_split() %>% {.[[1]]} -> df_i

      # get info
      i <- 
        df_i %>%
        dplyr::distinct(gene, sample)
      info <- t(i[1,]) %>% paste(collapse = '.')
      p[[info]] <<- list()

      # get number of celltypes for plot dims
      p[[info]]$n_cts <<- 
        df_i %>% 
        dplyr::filter(!is.na(celltype)) %>% 
        dplyr::pull(celltype) %>% 
        dplyr::n_distinct()

      # plot by sample and by celltype
      list(by_sample = list(lvl = 'drivers_samtools_depth', title_suffix = ''),
           by_celltype = list(lvl = 'drivers_samtools_depth_by_celltype', title_suffix = '_by_celltype')) %>%
        purrr::map(function(x){

          # plot all
          p[[info]][[paste0('all', x$title_suffix)]] <<-
            plot_all(
              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
              p_exon_coords = exon_coords %>% dplyr::filter(gene == i$gene),
              p_title = paste0(info, x$title_suffix)
            )

          # plot exons only
          p[[info]][[paste0('exons', x$title_suffix)]] <<-
            plot_exons(
              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
              p_title = paste0(info, x$title_suffix)
            )

        })

    })

>>>>

---
title: "Driver coverage"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  id: NULL
  depths: NULL
  drivers: NULL
  ref_cds: NULL
  cache_dir: './driver_coverage_cache/'
  rerun: true
---

@@

knitr::opts_knit$set(root.dir = './')
knitr::opts_chunk$set(
  warning = FALSE, 
  dpi = 300, 
  message = FALSE,
  cache.path = params$cache_dir)

@@

We read in drivers and their depths.

```{r read_cov}
# read coverage
cov <- xfun::cache_rds({
  cov <- 
    readr::read_tsv('depths.tsv') %>% 
    dplyr::mutate(chr = as.character(chr))
  cov
}, file = 'cov.rds', rerun = params$rerun)

# read drivers
drivers <- readr::read_tsv(params$drivers)
```

@@

```{r features}
# get ref_cds object from DNDSCV
ref_cds <- 
  readRDS(params$ref_cds)

# annotate gene features
cov <- xfun::cache_rds({

  # get all positions in each gene
  gene_spans <-
    ref_cds[c(drivers$gene)] %>%
    purrr::map(function(i) {
      tibble::tibble(
        chr = as.character(i$chr),
        start = i$intervals_cds[, 1],
        end = i$intervals_cds[, 2]) %>%
        dplyr::mutate(exon = paste0('exon', dplyr::row_number()))
    })

  # drop those not found
  gene_spans[is.na(names(gene_spans))] <- NULL

  # annotate exons
  exons <-
    gene_spans %>%
    dplyr::bind_rows(.id = 'gene') %>%
    dplyr::mutate(pos = purrr::map2(start, end, ~ .x:.y)) %>%
    tidyr::unnest(cols = 'pos') %>%
    dplyr::select(-start, -end)
  cov <-
    cov %>%
    dplyr::left_join(exons, by = c('chr', 'pos', 'gene')) %>%
    dplyr::mutate(exon = ifelse(is.na(exon), 'intron', exon),
                  feature = ifelse(exon == 'intron', 'intron', 'exon'))

@@

  cov %>% 
    dplyr::group_by(gene) %>%
    dplyr::group_split() %>%
    purrr::map(function(df_i) {

      # for testing:
      # cov %>% dplyr::group_by(gene) %>% dplyr::group_split() %>% {.[[1]]} -> df_i

      # get info
      gene_i <- unique(df_i$gene)
      p[[gene_i]] <<- list()

      # get number of celltypes for plot dims
      p[[gene_i]]$n_cts <<- 
        df_i %>% 
        dplyr::filter(!is.na(celltype)) %>% 
        dplyr::pull(celltype) %>% 
        dplyr::n_distinct()

      # plot by sample and by celltype
      df_i %>%
        dplyr::mutate(suffix = dplyr::case_when(is.na(celltype) ~ '',
                                                TRUE ~ '_by_celltype')) %>%
        {split(., .$suffix)} %>%
        purrr::map(function(lvl) {

          # get suffix
          p_suffix <- unique(lvl$suffix)

          # plot all
          p[[gene_i]][[paste0('all', p_suffix)]] <<-
            plot_all(
              p_df = lvl,
              p_exon_coords = exon_coords %>% dplyr::filter(gene == gene_i),
              p_title = paste0(gene_i, p_suffix)
            )

          # plot exons
          p[[gene_i]][[paste0('exons', p_suffix)]] <<-
            plot_exons(
              p_df = lvl,
              p_title = paste0(gene_i, p_suffix)
            )
        })
    })
